---
title: "Session 2 - Additional exercises"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, warning = FALSE, message = FALSE)
```


```{r}
library(tidyverse)

```


## Analysing quarterly UKgas

1. Read in "UKgas.csv" from the `./data` folder and inspect the data. (The data has been created from one of R datasets https://www.rdocumentation.org/packages/datasets/versions/3.6.2/topics/UKgas)

```{r}
UKgas <-  


```


2. Create a new tibble of the data in long format with a column to specify the quarter.

```{r}
UKgas_l <- 


```

3. Compute the mean quarterly UKgas consumption across years (Your new tibble will have four rows and 2 columns)

```{r}
UKgas_by_quarter <- 
  

```

4. Compute the mean gas consumption for each year (Your tibble will have 27 rows and 2 columns)

```{r}
UKgas_by_year <- 

```

5. **Bonus:** Convert your long tibble back to wide. This should be the same as the UKgas data. Compute the mean gas consumption by year. Have a look at https://stackoverflow.com/questions/50352735/calculate-the-mean-of-some-columns-using-dplyrmutate (Your tibble will have 27 rows and 6 columns). As you can see, working with long data is simpler in R.

```{r}
UKgas <- 

```


6. Plot the UKgas consumption by year as a line graph, with quarters shown in different colours. Change the axes labels to something of your choice and add a title.

```{r}

```

7. Plot the same as above, but include a line for the mean gas consumption across quarters. You will first need to append the UKgas_by_year to your data

```{r}

```

8. Create the same plot as above (including the mean), but use thin lines for quarter, and a thick line for the mean. You will need to add a new numeric variable to the data used in the previous exercise that specifies a value for line thickness. See the examples in `?geom_line` for details around specifying aesthetics for the line graph and how to do this by group. You will also need to look at `?scale_linewidth`

```{r}


```
