---
title: "Session 2 - Additional exercises"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, warning = FALSE, message = FALSE)
```


```{r}
library(tidyverse)

```


## Analysing quarterly UKgas

1. Read in "UKgas.csv" from the `./data` folder and inspect the data. (The data has been created from one of R datasets https://www.rdocumentation.org/packages/datasets/versions/3.6.2/topics/UKgas)

```{r}
UKgas <- read_csv("./data/UKgas.csv")
head(UKgas)

```


2. Create a new tibble of the data in long format with a column to specify the quarter.

```{r}
UKgas_l <- UKgas %>% 
  pivot_longer(cols = -year, 
               names_to = "quarter", 
               values_to = "gas_consumption")


```

3. Compute the mean quarterly UKgas consumption across years (Your new tibble will have four rows and 2 columns)

```{r}
UKgas_by_quarter <- UKgas_l %>% 
  group_by(quarter) %>% 
  summarise(mean_quarterly_gas = mean(gas_consumption, 
                                      na.rm = TRUE))


```

4. Compute the mean gas consumption for each year (Your tibble will have 27 rows and 2 columns)

```{r}
UKgas_by_year <- UKgas_l %>% 
  group_by(year) %>% 
  summarise(mean_annual_gas = mean(gas_consumption, 
                                      na.rm = TRUE))


```

5. **Bonus:** Convert your long tibble back to wide. This should be the same as the UKgas data. Compute the mean gas consumption by year. Have a look at https://stackoverflow.com/questions/50352735/calculate-the-mean-of-some-columns-using-dplyrmutate (Your tibble will have 27 rows and 6 columns). As you can see, working with long data is simpler in R.

```{r}
UKgas <- UKgas %>% 
  mutate(mean_annual_gas =rowMeans(select(., Qtr1,
                                  Qtr2,
                                  Qtr3, 
                                  Qtr4)))


```


6. Plot the UKgas consumption by year as a line graph, with quarters shown in different colours. Change the axes labels to something of your choice and add a title.

```{r}
g <- ggplot(UKgas_l, aes(x = year, 
                         y = gas_consumption, 
                         group = quarter,
                         colour = quarter)) + 
  geom_line() +
  xlab("Year") +                           ## Alternatively, all labels can be changed in a single line using labels(title = "", x="", y="")
  ylab("Gas consumption (mTherms)")

g
```

7. Plot the same as above, but include a line for the mean gas consumption across quarters. You will first need to append the UKgas_by_year to your data

```{r}
UKgas_l_with_mean <- UKgas_l %>% 
  bind_rows(UKgas_by_year %>% 
              # Specify the value that should appear in the "quarter" column
              mutate(quarter = "Mean") %>%
              # ensure that column names match
              rename(gas_consumption = mean_annual_gas))

g2 <- ggplot(UKgas_l_with_mean, aes(x = year, 
                         y = gas_consumption, 
                         group = quarter, 
                         colour = quarter)) + 
  geom_line() +
  xlab("Year") + 
  ylab("Gas consumption (mTherms)")

g2
```

8. Create the same plot as above (including the mean), but use thin lines for quarter, and a thick line for the mean. You will need to add a new numeric variable to the data used in the previous exercise that specifies a value for line thickness. See the examples in `?geom_line` for details around specifying aesthetics for the line graph and how to do this by group. You will also need to look at `?scale_linewidth`

```{r}
UKgas_l_with_mean <- UKgas_l_with_mean %>% 
  # Add a linewidth value. The numbers are not crucial as these will be rescaled in the plot
  mutate(linewidth = if_else(quarter == "Mean", 2, 1))

g3 <- ggplot(UKgas_l_with_mean, aes(x = year, 
                         y = gas_consumption, 
                         group = quarter, 
                         colour = quarter)) + 
  geom_line(aes(linewidth = linewidth)) +
  #specify the range that the linewidths should span, and disable the linewidth legend
  scale_linewidth(range=c(0.1, 2), guide=FALSE) +
  xlab("Year") + 
  ylab("Gas consumption (mTherms)")

g3

```
