---
title: "Session 3 Solutions"
author: ""
date: "08-10-2024"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```
## Overview

This session will aim to give you a chance to work through a problem that may be similar to one you would face your daily role. It will use functions we have covered in session 1 and 2, as well as some helpful new functions.

For the following coding challenges you'll use some aggregated data from the International Passenger Survey from 2011-2013. These data provide counts of visits to Wales each month by visit "type", e.g. business, holiday, etc. They were provided by ONS for our introductory SAS course.

Each section has an (optional) **stretch** challenge that may require functions not covered in the course so far. This is an opportunity to begin coding independently: searching for appropriate functions online, or using the help pages to understand ***hinted*** functions and how to use them.

## Section 1: Importing datasets

The data are in the "Data" folder.

### 1.1: Read in the CSV file `visits2011.csv`.


```{r import_csv}
# import library 
library(tidyverse)

visits_csv <- read_csv("Data/visits2011.csv")
```

### 1.2: Read in the Excel file `visits201213.xlsx`.

***Hint:*** use the `read_excel` function from the `readxl` package. Use the command `?read_excel` to view the help page.

```{r import_excel}
library(readxl)

visits_xl <- read_excel("Data/visits201213.xlsx")
```

### 1.3: Examine the two datasets. Whats the difference between these two datasets?

Depends on the functions used to read the datasets in. If all else is the same, the difference is the period covered.

### 1.4: Stretch: Read in TRNG.visits dataset from SAS.

If you have SAS installed on your computer, open SAS and use `proc contents` to find the file path of TRNG.ALL_VISITS. Note that the folder names may be different to that seen in the SAS server. The SAS libraries are located on the shared drive `"//s0177a/sasdata1/<path>/<filename>.sas7bdat"`. 

***Hint***: use `read_sas("//s0177a/sasdata1/<path>/<filename>.sas7bdat")` to read in the SAS datafile.


```{r import_SAS}

visits_sas <- haven::read_sas("//s0177a/sasdata1/Training/all_visits.sas7bdat") 
```

## Section 2: Manipulating data

### 2.1: Combine the datasets from 1.1 and 1.2.

Use functions from earlier in the course to combine the datasets.

```{r combine}
visits <- bind_rows(visits_csv,visits_xl)
```

### 2.2: Convert the dataset to a "long" (i.e. "tidy") format.

"tidy" format is a long data format with one measurement per row.

```{r pivot}
visits_long <- visits %>% 
  pivot_longer(-c("period", "source3"),
               names_to = "visit_type", 
               values_to = "no_of_trips")
```

### 2.3: Create new dataset using tibble to use as a lookup table to add the season corresponding to each month.

e.g. Winter: Dec, Jan, Feb/ Spring: Mar, Apr, May/ Summer: Jun, Jul, Aug/ Autumn: Sep, Oct, Nov

```{r lookup}
# Create the lookup
seasons <- tibble(month =seq(1,12), 
                  season = c(rep("winter", 2), rep("spring", 3),
                             rep("summer", 3), rep("autumn", 3), "winter"))

# Create the match variable
visits_months <- visits_long %>% 
  mutate(month = substr(period,5,6) %>%
  as.numeric())

# Combine the datasets
visits_seasons <- visits_months %>% 
  left_join(seasons, by = "month") %>% 
  # Optionally, remove the match variable as it isn't required
  select(-month)
```

### 2.4: Stretch
#### A: Create a `date` variable using the `period` variable. 
***Hint:*** Consider the function `as_date()` in the `lubridate` package. First you will need to modify the `period` variable into a form that `lubridate` can interpret as a date. 

#### B: Use the `date` variable to create a year variable called `calendar_year`.
***Hint:*** Use the `year()` function to create your `calendar_year` variable.

```{r dates}
library(lubridate)

# Create date variable
visits_date <- visits_seasons %>% 
  mutate(date = paste0(period,"01") %>% 
           as_date())

# Create year variable
visits_year <- visits_date %>% 
  mutate(calendar_year = year(date))
```

## Section 3: Analysing data

### 3.1: Calculate the total number of trips made during each season.

Create a new dataset containing the total number of trips for each of the 4 seasons.

```{r avg_visits_all}
visits_seasons_total <- visits_seasons %>%
  group_by(season) %>%
  summarise(total_trips_all = sum(no_of_trips)) 
```

### 3.2: For each visit type, find the total number of trips made during each season.

Create a new dataset containing the total number of trips made during the 4 seasons, for each of the visit types (business, holiday, etc.). ***Hint:*** You can group by more than one variable within tidyverse.

```{r avg_visits}
visits_types_seasons_total <- visits_seasons %>%
  group_by(visit_type, season) %>%
  summarise(total_trips = sum(no_of_trips)) 

```

### 3.3: For each each season, find the percentage of trips made for each of the visit types. Round this percentage to two decimal places.

```{r pct_visits}
visits_types_seasons_pct <- visits_types_seasons_total %>%
  full_join(visits_seasons_total,
            by = "season") %>%
  mutate(pct = total_trips / total_trips_all) %>% 
  mutate(pct = round(pct, digits = 2))
```

Alternatively using group_by

```{r pct_visits2}
visits_types_seasons_pct2 <- visits_seasons %>%
  group_by(visit_type, season) %>%
  summarise(no_of_trips = sum(no_of_trips)) %>%
  group_by(season) %>%
  mutate(pct = no_of_trips / sum(no_of_trips)) %>% 
  mutate(pct = round(pct, digits = 2))
```

### 3.4: Stretch: Calculate the monthly change in number of trips, by visit type.

Using `visits_date`, create a dataset containing the difference in number of trips compared to the previous month, for each visit type.

***Hint:*** The `lag()` function in the `dplyr` tidyverse package can be used to reference values in previous rows.

```{r diff_visits}
visits_diff <- visits_date %>%
  group_by(visit_type) %>%
  arrange(date) %>%
  mutate(diff = no_of_trips - lag(no_of_trips, default = NA_real_))
```


## Section 4: Creating charts using ggplot

### 4.1. Create a bar chart showing the number of trips by calendar year and season.

***Hint:*** Use the `visits_year` data you created in 2.4. Consider carefully the x and y variables in `aes()`.

```{r output_chart_1}
# Creating a basic bar chart using ggplot
# Choosing the x and y variables
bar_graph_trips <- ggplot(data = visits_year,
                          aes(x = calendar_year,
                              y = no_of_trips)) +
  
  # Define chart type and colour bars by season
  geom_col(aes(fill = season), position = "dodge")

bar_graph_trips
```

### 4.2 Export `visits_date` to an Excel workbook named "Ex 4.2" and save your bar chart as a jpg.

***Hint:*** Use the function `write_xlsx()` from the package `writexl`. Use the function `ggsave()` to save your bar chart.

``` {r export}
library(writexl)

# Save table
write_xlsx(visits_date, "Ex 4.2.xlsx")

# Save chart
ggsave("bar_chart.jpg", plot = bar_graph_trips)
```

### 4.3. **Stretch:** Amend `bar_graph_trips` to create a publication standard chart.

Some suggested improvements:

* Order the seasons (***hint:*** do this before you start plotting using the `levels` argument in the `factor()` function)
* Re-label the x and y axis (***hint:*** "Labels and Legends" section of the ggplot cheat sheet)
* Set the theme to show a white background with horizontal gridlines (***hint:*** "Themes" section of the ggplot cheat sheet)
* Change the colour of the bars
* Format the y-axis label with thousands comma (***hint:*** use the `scales` package)
* Define the y-axis limits and breaks

``` {r output_chart_2}
# Order the seasons
visits_year <- visits_year %>% 
  mutate(season = factor(season, levels = c("spring", "summer", 
                                            "autumn", "winter")))

# Use sgplot, an R package for creating accessible plots 
# in the Scottish Government
sgplot::use_sgplot()

# Create chart
bar_graph_trips <- ggplot(data = visits_year,
                          aes(x = calendar_year,
                              y = no_of_trips)) +
  
  # Define chart type and colour bars by season
  geom_col(aes(fill = season), position = "dodge") +
  
  # Label the axes
  labs(x = "Calendar year",
       y = "Number of trips per year",
       
       # These lines add a title and subtitle to the chart, 
       # however it's better for accessibility if titles are 
       # not embedded in charts.
       title = "Visits to Wales by season, 2011 to 2013",
       subtitle = "Source: International Passenger Survey") + 
  
  # Format y-axis labels with thousands comma and define limits
  scale_y_continuous(labels = scales::comma,
                     limits = c(min = 0, max = 6000),
                     breaks = seq(0, 6000, by = 1000)) +
  
  # Capitalise legend title and text
  scale_fill_discrete(name = "Season",
                      labels = c("Spring", "Summer", "Autumn", "Winter")) +
  
  # The sgplot theme sets the axis text as horizontal.
  # This sets the y-axis title back to default (a 90 degree angle)
  # if preferred.
  theme(axis.title.y = element_text(angle = 90))

bar_graph_trips
```

