---
title: "Intro to R session 3"
author: ""
date: "March 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(DT)
library(tidyverse)
```


# Background

* Social Security Scotland administers four benefits. The types of cases processed for each of the benefits can be New Cases, Change of Circumstances or Award Reviews.
* The data available is the forecasted number of cases for each of the types for each of the four benefits
* Clients receive the decision on their case 2 months after applying
* Clients who are unhappy with the decision made on their case can apply for a review of the case. They receive a decision after 1 month. If they are still unhappy, they can appeal the decision
* SSS requires a forecast of FTE to process the reviews and appeals 

# Client Output requirements
* A plot of the total FTE over time
* A table with the annual FTE to process reviews and appeal defined as the median 

# Assumptions



```{r Assumptions_Redeterminations, echo=FALSE}

Benefits = c("Benefit1", "Benefit2", "Benefit3", "Benefit4") 
Types = c('Award Reviews', 'Change of Circ', 'New Cases' )
review_prop <- dplyr::tibble(type = Types, 
                                       Benefit1 = c(0.1, 0, 0.2), 
                                       Benefit2 = c(0.5, 0.8, 0.1), 
                                       Benefit3 = c(0, 0, 0), 
                                       Benefit4 = c(1, 1, 1)
                                       ) 
knitr::kable(review_prop, caption = "Proportion of total cases that require review" )
```




```{r Assumptions_appeals, echo = FALSE}
appeals_prop <- dplyr::tibble(type = Types, 
                                       Benefit1 = c(0.1, 0.5, 0.6), 
                                       Benefit2 = c(0.5, 1, 1), 
                                       Benefit3 = c(0, 0, 0), 
                                       Benefit4 = c(0, 0, 0)
)
knitr::kable(appeals_prop,  caption = "Proportion of reviews that appeal")
```

```{r Assumptions_redetermination_times, echo = FALSE}
review_time <- dplyr::tibble(type = Types, 
                                       Benefit1 = c(10, 10, 15), 
                                       Benefit2 = c(10, 12, 8), 
                                       Benefit3 = c(0, 0, 0), 
                                       Benefit4 = c(10, 15, 5)
)
knitr::kable(review_time, caption = "Time required to process reviews (min)")
```


```{r Assumptions_appeals_times, echo = FALSE}
appeals_time <- dplyr::tibble(type = Types, 
                                       Benefit1 = c(10, 10, 15), 
                                       Benefit2 = c(10, 12, 8), 
                                       Benefit3 = c(0, 0, 0), 
                                       Benefit4 = c(10, 15, 5)
)
knitr::kable(appeals_time, caption = "Time required to process appeals (min)")
```

* The client has estimated that out of a `r 7.4*60` min day, a person is productive for 75% of the time

# Data

* Benefits1.csv, Benefits2.csv, Benefits3.csv, Benefits4.csv contain the forecasted number of applications of each type for the period of interest
* working_days.csv contains the number of working days in each month over the period of interest