---
title: "session3_revised"
author: ""
date: "2024-03-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Overview

This session will aim to give you a chance to work through a problem that may be similar to one you would face your daily role. It will use functions we have covered in session 1 and 2, as well as some helpful new functions.

## Section 1: Importing dataset

1.  Read in the CSV file `"visits2011.csv"`
2.  Read in the Excel file `"visits201213.xlsx"`
3.  Examine the two datasets - what differs about the outputs of each method?
4.  **Stretch**: Read in the SAS dataset TRNG.ALL_VISITS to R

Some hints in the sections below....

### 1.1: Read in visits2011.csv

The data are in the folder `"intro_to_r/session3/Data"`

### 1.1: Read in visits2011.csv

The data are in the folder `"intro_to_r/session3/Data"`

### 1.2: Read in visits201213.xlsx

Hint: use the `readxl` package. Use the command `??readxl` to view the help page for the package

```{r import_excel}
#import library 
library(readxl)

visits_xl <- 

```

### 1.3: Whats the difference between these two datasets?

Use base R functions to look at the structure, values and types of variables in each dataset.

### 1.4: Read in TRNG.ALL_VISITS dataset from SAS.

Hint: use the function `haven::read_sas` to read in SAS datafiles.

The SAS libraries are located on the shared drive `"//s0177a/sasdata1/<path>/<filename>.sas7bdat"`. Note that the folder names may be different to that seen in the SAS server.

You can use proc contents in SAS to find the exact file path.

```{r import_SAS}
library(haven)

all_visits_sas <- 
```

## Section 2: Manipulating data

1.  Combine the datasets from 1.1 and 1.2 as appropriate. Use the resulting data (or the dataset from 1.4) for the following tasks:
2.  Convert the dataset to a "long" (i.e. "tidy") format
3.  Create a "season" variable in the dataset by building a lookup table to match a season value to each value of month
4.  **Stretch**: Create a date format variable using information in the "period" column

Some hints in the sections below....

### 2.1: Combine datasets

Use functions from earlier in the course to combine the datasets

### 2.2: Pivot data into tidy format

"tidy" format is a long data format with one measurement per row.

```{r pivot}
all_visits_sas_long <-
```

### 2.3: Create new dataset using tibble so use as a lookup table to add the season corresponding to each month.

Create a table that corresponds specific months to season, e.g. Winter: Dec, Jan, Feb/ Spring: March, April, May/ Summer: June, July, August/ Autumn: Sep, Oct, nov

Match this information in by extracting the month information from the "period" variable: consider the functions `substr` and `as.numeric`

```{r lookup}
seasons <- 
```

### 2.4: Stretch: Create a date variable using the "period" column

Consider the function `as_date()` in the `lubridate` package. First you will need to modify the period variable into a form that `lubridate` can interpret as a date

```{r dates}
library(tidyverse)
library(lubridate)



```

## Section 3: Analysing data

1.  Calculate the total number of trips made during each season
2.  For each visit type, find the total number of trips made during each season
3.  For each each season, find the percentage of trips made for each of the visit types
4.  **Stretch**: Calculate the monthly change in number of trips, by visit type

### 3.1: Calculate the total number of trips made during each season

Create a new dataset summarising the average number of trips for each of the 4 seasons.

### 3.2: For each visit type, find the total number of trips made during each season

Create a new dataset containing the average number of trips made during the 4 seasons, for each of the visit types (business, holiday, etc.). You can group by more than one variable within tidyverse

### 3.3: For each each season, find the percentage of trips made for each of the visit types

Consider joining the two datasets created in 3.1 and 3.2,

### 3.4: Stretch: Calculate the monthly change in number of trips, by visit type

Create a dataset containing the difference in number of trips compared to the previous month, for each visit type. The `lag()` function in the `dplyr` tidyverse package can be used to reference values in previous rows
