---
title: "session3_revised"
author: ""
date: "2024-03-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Overview

This session will aim to give you a chance to work through a problem that may be similar to one you would face your daily role. It will use functions we have covered in session 1 and 2, as well as some helpful new functions.

## Section 1: Importing datasets

1.  Read in the CSV file `"visits2011.csv"`
2.  Read in the Excel file `"visits201213.xlsx"`
3.  Examine the two datasets - what differs about the outputs of each method?
4.  **Stretch**: Read in the SAS dataset TRNG.ALL_VISITS to R

Some hints in the sections below....

### 1.1: Read in visits2011.csv

The data are in the folder `"intro_to_r/session3/Data"`

### 1.1: Read in visits2011.csv

The data are in the folder `"intro_to_r/session3/Data"`

### 1.2: Read in visits201213.xlsx

Hint: use the `readxl` package. Use the command `??readxl` to view the help page for the package

```{r import_excel}
#import library 
library(readxl)

visits_xl <- 

```

### 1.3: Whats the difference between these two datasets?

Use base R functions to look at the structure, values and types of variables in each dataset.

### 1.4: Read in TRNG.ALL_VISITS dataset from SAS.

Hint: use the function `haven::read_sas` to read in SAS datafiles.

The SAS libraries are located on the shared drive `"//s0177a/sasdata1/<path>/<filename>.sas7bdat"`. Note that the folder names may be different to that seen in the SAS server.

You can use proc contents in SAS to find the exact file path.

```{r import_SAS}
library(haven)

all_visits_sas <- 
```

## Section 2: Manipulating data

1.  Combine the datasets from 1.1 and 1.2 as appropriate. Use the resulting data (or the dataset from 1.4) for the following tasks.
2.  Convert the dataset to a "long" (i.e. "tidy") format
3.  Create a "season" variable in the dataset by building a lookup table to match a season value to each value of month
4.  **Stretch**: Create a date format variable using information in the "period" column

Some hints in the sections below....

### 2.1: Combine datasets

Use functions from earlier in the course to combine the datasets

### 2.2: Pivot data into tidy format

"tidy" format is a long data format with one measurement per row.

```{r pivot}
all_visits_sas_long <-
```

### 2.3: Create a new dataset using tibble to use as a lookup table to add the season corresponding to each month.

Create a table that corresponds specific months to season, e.g. Winter: Dec, Jan, Feb/ Spring: March, April, May/ Summer: June, July, August/ Autumn: Sep, Oct, nov

Match this information in by extracting the month information from the "period" variable: consider the functions `substr` and `as.numeric`

```{r lookup}
seasons <- 
```

### 2.4: Stretch: Create a date variable using the "period" column

Consider the function `as_date()` in the `lubridate` package. First you will need to modify the period variable into a form that `lubridate` can interpret as a date

```{r dates}
library(tidyverse)
library(lubridate)



```

## Section 3: Analysing data

1.  Calculate the total number of trips made during each season
2.  For each visit type, find the total number of trips made during each season
3.  For each each season, find the percentage of trips made for each of the visit types
4.  **Stretch**: Calculate the monthly change in number of trips, by visit type

### 3.1: Calculate the total number of trips made during each season

Create a new dataset summarising the average number of trips for each of the 4 seasons.

### 3.2: For each visit type, find the total number of trips made during each season

Create a new dataset containing the average number of trips made during the 4 seasons, for each of the visit types (business, holiday, etc.). You can group by more than one variable within tidyverse

### 3.3: For each each season, find the percentage of trips made for each of the visit types

Consider joining the two datasets created in 3.1 and 3.2,

### 3.4: Stretch: Calculate the monthly change in number of trips, by visit type

Create a dataset containing the difference in number of trips compared to the previous month, for each visit type. The `lag()` function in the `dplyr` tidyverse package can be used to reference values in previous rows

## Section 4: Creating output tables

1.  Create an output table showing the median number of trips by calendar year
2.  Create an output table showing the number of trips as a percentage of all trips taken by calendar year 3 . **Stretch**: Modify the table created above to show the percentages rounded to the nearest whole number

### 4.1: Create an output table showing the median and total number of trips by calendar year.

Consider using `year()` in the `lubridate` package, once the date variable has been created (see 2.4), or extract the year as a substring of \`source3\`.

Tidy the column headings using `rename()`

### 4.2: Create an output table showing the number of trips as a percentage of all trips taken by calendar year

Consider using `ungroup()` in the `tidyverse` package after calculating the number of trips per year. Use `select()` to choose the columns needed for the output table.

### 4.3: Stretch: Modify the table created above to show the percentages rounded to the nearest whole number

Think of ways to adjust the code used to create the table in 4.2. The `round()` function can be used to calculate rounded figures.

## Section 5: Creating charts using ggplot

1.  Create a bar chart showing the number of trips by calendar year
2.  Create a facet bar plot for each season showing the number of trips by calendar year
3.  **Stretch**: Create a stacked bar chart showing the number of trips by calendar year and season

### 5.1: Create a bar chart showing the number of trips by calendar year

Consider using `year()` in the `lubridate` package, once the date variable has been created (see 2.4), or extract the year as a substring of \`source3\`.

Consider carefully the x and y variables in `aes()`.

### 5.2: Create a facet bar plot for each season showing the number of trips by calendar year

Consider carefully the variables in `aes()`, you should be including more that just x and y variables here. Use `facet_grid()` to create the facet bar plot.

### 5.3: Stretch: Create a stacked bar chart showing the number of trips by calendar year and season

Think about what should be included in `geom_col()` to specify that the bars should be stacked.
