---
title: "session3_resvised_solutions"
author: ""
date: "2024-03-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Overview

This session will aim to give you a chance to work through a problem that may be similar to one you would face your daily role. It will use functions we have covered in session 1 and 2, as well as some helpful new functions. 

## Section 1: Importing dataset 

### 1.1: Read in TRNG.ALL_VISITS dataset from SAS. 

Hint: use the following code.... haven::read_sas("//s0177a/sasdata1/folder_names/dataset_name.sas7bdat"). Note that the folder names may be different to that seen in the SAS server. Use proc contents in SAS to find the exact file path. 

```{r import_SAS}

all_visits_sas <- haven::read_sas("//s0177a/sasdata1/
```

### 1.2: Read in TRNG.ALL_VISITS dataset as xlsx. 

Hint: use ?readxl help page

```{r import_excel}
#import library 
library(readxl)

all_visits_xl <- 

```

### 1.3: Whats the difference between these two datasets?


## Section 2: Working with dates

### 2.1: Using the all_visits dataset imported from SAS. Change the period column into a date 

```{r dates}
library(tidyverse)
library(lubridate)



```


### 2.2: Pivot data into tidyverse format

```{r pivot}
all_visits_sas_long <-
```

### 2.3: Create new dataset using tibble so use as a lookup table to add the season correspinding to each month. 

e.g. Winter: Dec, Jan, Feb/ Spring: March, April, May/ Summer: June, July, August/ Autumn: Sep, Oct, nov


```{r lookup}
seasons <- 
```


