---
title: "session3_resvised_solutions"
author: ""
date: "2024-03-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Overview

This session will aim to give you a chance to work through a problem that may be similar to one you would face your daily role. It will use functions we have covered in session 1 and 2, as well as some helpful new functions.

## Section 1: Importing dataset

### 1.1: Read in visits2011.csv

The data are in the folder "intro_to_r/session3/Data"

```{r import_csv}
library(tidyverse)

visits_csv <- read_csv("Data/visits2011.csv")

```

### 1.2: Read in visits201213.xlsx

Hint: use the readxl package. Use the command ?readxl to view the help page

```{r import_excel}
#import library 
library(readxl)

visits_xl <- read_excel("Data/visits201213.xlsx")

```

### 1.3: Whats the difference between these two datasets?

Depends on the functions used to read the datasets in. If all else is the same, the difference is the period covered.

### 1.4: Stretch: Read in TRNG.ALL_VISITS dataset from SAS.

Hint: use the function read_sas in the haven package to read in SAS datafiles.

The SAS libraries are located on the shared drive "//s0177a/sasdata1/<path>/<filename>.sas7bdat". Note that the folder names may be different to that seen in the SAS server.

You can use proc contents in SAS to find the exact file path.

```{r import_SAS}

all_visits_sas <- haven::read_sas("//s0177a/sasdata1/training/all_visits.sas7bdat") 
```

## Section 2: Manipulating data

### 2.1: Combine datasets

Use functions from earlier in the course to combine the datasets

```{r combine}

all_visits = bind_rows(visits_csv,visits_xl) %>% glimpse()

```

### 2.2: Pivot data into tidy format

"tidy" format is a long data format with one measurement per row.

```{r pivot}
all_visits_long <- all_visits %>% 
  pivot_longer(-c("period", "source3"),
               names_to= "holiday_type", 
               values_to = "no_of_trips") %>% glimpse()
```

### 2.3: Create new dataset using tibble so use as a lookup table to add the season corresponding to each month.

e.g. Winter: Dec, Jan, Feb/ Spring: March, April, May/ Summer: June, July, August/ Autumn: Sep, Oct, nov

```{r lookup}
# Create the lookup
seasons <- tibble(month =seq(1,12), 
                  season = c(rep("winter", 2), rep("spring", 3),rep("summer", 3), rep("autumn", 3),"winter"))

# Create the match variable
all_visits_months <- all_visits_long %>% 
  mutate(month = substr(period,5,6) %>% as.numeric()) %>% glimpse()

# Combine the datasets
all_visits_seasons <- all_visits_months %>% 
  left_join(seasons, by = "month") %>% 
  # Optionally, remove the match variable as it isn't required
  select(-month) %>% glimpse()

```

### 2.4: Stretch: Create a date variable using the "period" column

```{r dates}
library(tidyverse)
library(lubridate)

all_visits_date <- all_visits_seasons %>% 
  mutate(date = paste0(period,"01") %>% 
           as_date()) %>% glimpse()

```

## Section 3: Analysing data

### 3.1: Calculate the total number of trips made during each season

```{r avg_visits_all}
library(knitr)

all_visits_seasons_total <- all_visits_seasons %>%
  group_by(season) %>%
  summarise(total_trips_all = sum(no_of_trips)) #%>% glimpse()

kable(all_visits_seasons_total)
```

### 3.2: For each visit type, find the total number of trips made during each season

```{r avg_visits}
visits_types_seasons_total <- all_visits_seasons %>%
  group_by(holiday_type, season) %>%
  summarise(total_trips = sum(no_of_trips)) #%>% glimpse()

kable(visits_types_seasons_total)
```

### 3.3: For each each season, find the percentage of trips made for each of the visit types

```{r pct_visits}
visits_types_seasons_pct <- visits_types_seasons_total %>%
  full_join(all_visits_seasons_total,
            by = "season") %>%
  mutate(pct = total_trips / total_trips_all) #%>% glimpse

kable(visits_types_seasons_pct)
```

alternatively using group_by-mutate

```{r pct_visits2}
visits_types_seasons_pct2 <- all_visits_seasons %>%
  group_by(holiday_type, season) %>%
  summarise(no_of_trips = sum(no_of_trips)) %>%
  group_by(season) %>%
  mutate(pct = no_of_trips / sum(no_of_trips)) %>% glimpse

kable(visits_types_seasons_pct2)
```

### 3.4: Stretch: Calculate the monthly change in number of trips, by visit type

```{r diff_visits}

all_visits_diff <- all_visits_date %>%
  group_by(holiday_type) %>%
  arrange(date) %>%
  mutate(diff = no_of_trips - lag(no_of_trips, default = NA_real_)) #%>% glimpse()

kable(all_visits_diff)

```

## Section 4: Creating output tables

### 4.1: Create an output table showing the median number of trips by calendar year.

Consider using `year()` in the `lubridate` package, once the date variable has been created (see 2.4). Tidy the column headings using `rename()`

```{r output_table_1}

all_visits_output <- all_visits_date %>% 
  
  # Creating the calendar year column and grouping by this new variable
  mutate(calendar_year = lubridate::year(date)) %>% 
  group_by(calendar_year) %>% 
  
  # Calculating the median and total number of trips
  summarise(median_trips = median(no_of_trips),
            total_trips = sum(no_of_trips)) %>%
  
  # Renaming columns in the output table to remove R syntax
  rename("Calendar Year" = calendar_year, 
         "Median number of trips per year" = median_trips,
         "Total number of trips per year" = total_trips)

kable(all_visits_output)

```


### 4.2: Create an output table showing the number of trips as a percentage of all trips taken by calendar year

Consider using `ungroup()` in the `tidyverse` package after calculating the number of trips per year. Use `select()` to choose the columns needed for the output table.

```{r output_table_2}

all_visits_output_pct <- all_visits_date %>%
  
  # Creating the calendar year column and grouping by this new variable
  mutate(calendar_year = year(date)) %>%
  group_by(calendar_year) %>%
  
  # Calculating the total number of trips in each calendar year and then un-grouping the data
  summarise(total_trips = sum(no_of_trips)) %>%
  ungroup() %>%
  
  # Calculate the grand total of trips across all years
  mutate(grand_total_trips = sum(total_trips)) %>%
  
  # Calculate the percentage of total trips per year by the grand total
  
  mutate(percentage_of_total_trips = (total_trips / grand_total_trips) * 100) %>%
  
  # Select the columns as needed for the final output table
  select(calendar_year, 
         total_trips, 
         percentage_of_total_trips) %>%
  
  # Renaming columns in the output table to remove R syntax
  rename("Calendar Year" = calendar_year, 
         "Total number of trips per year" = total_trips,
         "Percentage of total trips taken (%)" = percentage_of_total_trips)

kable(all_visits_output_pct)

```


### 4.3: Stretch: Modify the table created above to show the percentages rounded to the nearest whole number

Think of ways to adjust the code used to create the table in 4.2. The `round()` function can be used to calculate rounded figures.

```{r output_table_3}

all_visits_output_pct_rounded <- all_visits_date %>%
  
  # Creating the calendar year column and grouping by this new variable
  mutate(calendar_year = year(date)) %>%
  group_by(calendar_year) %>%
  
  # Calculating the total number of trips in each calendar year and then un-grouping the data
  summarise(total_trips = sum(no_of_trips)) %>%
  ungroup() %>%
  
  # Calculate the grand total of trips across all years
  mutate(grand_total_trips = sum(total_trips)) %>%
  
  # Calculate the percentage of total trips per year by the grand total and rounding this figure
  mutate(percentage_of_total_trips = (total_trips / grand_total_trips) * 100,
         rounded_percentage_trips = round(percentage_of_total_trips)) %>%
  
  # Select the columns as needed for the final output table
  select(calendar_year, 
         total_trips, 
         percentage_of_total_trips, 
         rounded_percentage_trips) %>%
  
  # Renaming columns in the output table to remove R syntax
  rename("Calendar Year" = calendar_year, 
         "Total number of trips per year" = total_trips,
         "Percentage of total trips taken (%)" = percentage_of_total_trips,
         "Rounded Percentage of total trips taken (%)" = rounded_percentage_trips)

kable(all_visits_output_pct_rounded)

```


## Section 5: Creating charts using ggplot

### 5.1. Create a bar chart showing the number of trips by calendar year

Consider using `year()` in the `lubridate` package, once the date variable has been created (see 2.4). Consider carefully the x and y variables in `aes()`.

```{r output_chart_1}

# Creating a new dataset with the calendar year column added
all_visits_year <- all_visits_date %>% 
  mutate(calendar_year = lubridate::year(date))

# Creating a basic bar chart using ggplot
# Choosing the x and y variables
bar_graph_trips <- ggplot(data = all_visits_year,
                          aes(x = calendar_year,
                              y = no_of_trips)) +
  # Choosing a bar chart
  geom_col() +
  
  # Labelling the x axis
  xlab("Calendar Year") +
  
  # Labelling the y axis
  ylab("Number of trips per year") 

bar_graph_trips

```

### 5.2. Create a facet bar plot for each season showing the number of trips by calendar year

Consider carefully the variables in `aes()`, you should be including more that just x and y variables here. Use `facet_grid()` to create the facet bar plot.

```{r output_chart_2}

# Creating a new dataset with the calendar year column added
all_visits_year <- all_visits_date %>% 
  mutate(calendar_year = lubridate::year(date))

# Creating a facet bar plot 
# Choosing the x and y variables and grouping by season
bar_graph_trips_facet <- ggplot(data = all_visits_year,
                                aes(x = calendar_year,
                                    y = no_of_trips,
                                    group = season,
                                    colour = season)) +
  
  # Creating a bar chart and using colour to demonstrate the different seasons
  geom_col(aes(fill = season)) +
  
  # Making the bar chart into a facet plot for each season
  facet_grid(rows = vars(season)) +
  
  # Labelling the x axis
  xlab("Calendar Year") +
  
  # Labelling the y axis
  ylab("Number of trips per year") +
  
  # Setting the theme to show a white background with gridlines
  theme_bw()
 
bar_graph_trips_facet

```

### 5.3. Stretch: Create a stacked bar chart showing the number of trips by calendar year and season

Think about what should be included in `geom_col()` to specify that the bars should be stacked. 

```{r output_chart_3}

# Creating a new dataset with the calendar year column added
all_visits_year <- all_visits_date %>% 
  mutate(calendar_year = lubridate::year(date))

# Creating a stacked bar chart 
# # Choosing the x and y variables
bar_graph_trips_stacked <- ggplot(data = all_visits_year, 
                                            aes(x = calendar_year, 
                                                y = no_of_trips)) +
  
  # Creating a bar chart and using colour to demonstrate the different seasons
  geom_col(aes(fill = season), 
           position = "stack") + # Stacking the bars by season
  
  # Labelling the x axis
  xlab("Calendar Year") +
  
  # Labelling the y axis
  ylab("Number of trips per year") +
  
  # Labelling the legend for season
  labs(fill = "Season") + 
  theme_bw()
 
bar_graph_trips_stacked

```